---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
setwd("/Users/soobrosa/RepData_PeerAssessment1")
data <- read.csv(unz("activity.zip", "activity.csv"))
```

I will use **dplyr** and **gplot2** throughout this analysis.
```{r}
library(dplyr)
library(ggplot2)
```

## What is mean total number of steps taken per day?

Aggregate and plot:
```{r}
totals <- data %>%
  group_by(date) %>%
  summarise(total = sum(steps, na.rm = TRUE))

ggplot(totals, aes(x=total)) + geom_histogram(colour="black", fill="white")
```

The mean of the totals is:
```{r}
mean(totals$total)
```
The median of the totals is:
```{r}
median(totals$total)
```
## What is the average daily activity pattern?

Aggregate and plot:
```{r}
averages <- data %>%
  group_by(interval) %>%
  summarise(average = mean(steps, na.rm = TRUE))

ggplot(averages, aes(interval, average)) + geom_line()
```

The 5-minute interval containing the maximum number of steps is:
```{r}
averages[averages$average == max(averages$average), ]['interval']
```

## Imputing missing values

The total number of missing values in the dataset is:
```{r}
sum(!complete.cases(data)) 
```

My strategy for filling in all of the missing values in the dataset is using the the mean for the 5-minute interval.

Creating the new dataset
```{r}
data$average <- inner_join(data, averages)[, 4]

imputed <- data %>%
  mutate(steps = ifelse(is.na(steps), average, steps)) %>%
  select(-average)
```

Aggregate and plot:
```{r}
filled <- imputed %>%
  group_by(date) %>%
  summarise(total = sum(steps, na.rm = TRUE))

ggplot(filled, aes(x=total)) + geom_histogram(colour="black", fill="white")
```

The mean of the totals is:
```{r}
mean(filled$total)
```
The median of the totals is:
```{r}
median(filled$total)
```

Imputing distorted the measurements upwards and crammed mean and median together and made the total daily number of steps higher.

## Are there differences in activity patterns between weekdays and weekends?

Add a new column that differentiates between weekdays and weekends:
```{r}
imputed$date <- as.Date(imputed$date)
workweek <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
imputed$weekpart <-factor((weekdays(imputed$date) %in% workweek)+1L, levels=1:2, labels=c('weekend', 'weekday'))
```

Aggregate and plot:
```{r}
comparison <- imputed %>%
  group_by(weekpart, interval) %>%
  summarise(average = mean(steps, na.rm = TRUE))

ggplot(comparison, aes(interval, average )) + geom_line() + facet_wrap( ~ weekpart, ncol = 1)
```

The steps taken on weekends are more even and generally more in absolute terms.