---
title: "Statistical Inference Course Project - First Part by Daniel Molnar"
output: html_document
---

Overview
=

The first part of the Statistical Inference Course Project is a simulation exercise. In this project I will investigate the exponential distribution in R and compare it with the Central Limit Theorem with the distribution of averages of 40 exponentials.

Simulations
=

The exponential distribution can be simulated in R with `rexp(n, lambda)` where lambda is the rate parameter. The mean of exponential distribution is `1/lambda` and the standard deviation is also `1/lambda`. I set `lambda = 0.2` for all of the simulations. I will investigate the distribution of averages of 40 exponential with a thousand simulations.

```{r}
# set seed for reproducability
set.seed(642531)

# defaults
lambda <- 0.2
samples <- 40
simulations <- 1000

# here we go with the simulation
simulated_exponentials <- replicate(simulations, rexp(samples, lambda))
```

Sample Mean versus Theoretical Mean
=

I show the sample mean and compare it to the theoretical mean of the distribution.

```{r}
# calculate the means of from the simulation
simulation_means <- apply(simulated_exponentials, 2, mean)

# calculate the sample mean
sample_mean <- mean(simulation_means)
sample_mean

# calculate the theoretical mean
theoretical_mean <- 1/lambda
theoretical_mean
```

The sample mean is 5.001233 while the theoretical mean is 5. Figure 1 shows that the blue theoretical mean technically hides the green sample mean on the figure they are so close.

Sample Variance versus Theoretical Variance
=

I show how variable the sample is and compare it to the theoretical variance of the distribution.

```{r}
# calculate the variance of the sample
simulation_standard_deviation <- sd(simulation_means)
simulation_variance <- simulation_standard_deviation^2

# calculate the theoretical variance 
theoretical_standard_deviation <- (1/lambda)/sqrt(samples)
theoretical_variance <- ((1/lambda)*(1/sqrt(samples)))^2

simulation_standard_deviation

theoretical_standard_deviation

simulation_variance

theoretical_variance
```

The variance of the simulation is 0.5836537 while the theoretical variance is 0.625. The standard deviation of the simulation is 0.7639723 while the theoretical standard deviation is 0.7905694. 

Distribution
=

I show that the distribution is approximately normal.

Figure 2 shows that the normal distribution is pretty close to the histogram.

The almost straight line shows on figure 3 shows that due to the central limit theorem the distribution is close to a normal distribution.

\newpage

Appendix
=

Figure 1
==

```{r, out.width = '600px', out.height = '600px'}
# show them on a figure
library(ggplot2)
p <- ggplot(data = data.frame(simulation_means), aes(simulation_means)) + geom_histogram(colour="black", fill="white", binwidth = .2)                                                                                     
p + geom_vline(xintercept = sample_mean, colour = "green") + geom_vline(xintercept = theoretical_mean, colour = "blue") + ggtitle("Distribution of Means in the Simulations") + xlab("means")
```

Figure 2
==

```{r, out.width = '600px', out.height = '600px'}
p <- ggplot(data = data.frame(simulation_means), aes(x = simulation_means)) + geom_histogram(aes(y = ..density..), colour="black", fill="white", binwidth = .2)                                                                                     
p + stat_function(fun = dnorm, arg = list(mean = theoretical_mean, sd = theoretical_standard_deviation)) + ggtitle("Simulation versus Theory") + xlab("means")
```

Figure 3
==

```{r, out.width = '600px', out.height = '600px'}
qplot(sample = simulation_means) + geom_abline(intercept=5, slope=1, alpha=0.5)
```